---
title: "P8105_hw2_sri2121"
author: "Sameera"
date: "2026-10-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Loading the CSV file

```{r}
subway_df = 
  read_csv(file = "NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) |>
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "NO" ~ 0,
      "YES" ~ 1,
    )
  )
```

Questions:

1. How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.

```{r}
subway_df |> 
  select(station_name, line) |> 
  distinct()
```


2. How many stations are ADA compliant?

```{r}
subway_df |> 
  filter(ada == "TRUE") |> 
  select(station_name, line) |> 
  distinct()
```


3. What proportion of station entrances / exits without vending allow entrance?

```{r}
subway_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

Reformat data so that route number and route name are distinct variables. 
How many distinct stations serve the A train? 
Of the stations that serve the A train, how many are ADA compliant?

```{r}
subway_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A") |> 
  select(station_name, line) |> 
  distinct()

subway_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A", ada == TRUE) |> 
  select(station_name, line) |> 
  distinct()
```

















