---
title: "P8105_hw2_sri2121"
author: "Sameera"
date: "2026-10-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading libraries

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Loading the CSV file

```{r}
subway_df = 
  read_csv(file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) |>
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry, 
      "NO" ~ 0,
      "YES" ~ 1,
    )
  )
```

Questions:

1. How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.

```{r}
subway_df |> 
  select(station_name, line) |> 
  distinct()
```


2. How many stations are ADA compliant?

```{r}
subway_df |> 
  filter(ada == "TRUE") |> 
  select(station_name, line) |> 
  distinct()
```


3. What proportion of station entrances / exits without vending allow entrance?

```{r}
subway_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

Reformat data so that route number and route name are distinct variables. 
How many distinct stations serve the A train? 
Of the stations that serve the A train, how many are ADA compliant?

```{r}
subway_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A") |> 
  select(station_name, line) |> 
  distinct()

subway_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A", ada == TRUE) |> 
  select(station_name, line) |> 
  distinct()
```

## Problem 2

Importing, cleaning and combining data

```{r}
mrtrashwheel_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Mr. Trash Wheel",
                      skip = 1,
                      range = cell_cols("A:N")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(sports_balls = as.integer(round(sports_balls))) |> 
  mutate(trash_wheel = "MrTrashWheel") |> 
  dplyr::relocate(trash_wheel) |> 
  mutate(year = as.integer(year))

proftrashwheel_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Professor Trash Wheel",
                      skip = 1,
                      range = cell_cols("A:M")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "ProfTrashWheel") |> 
  dplyr::relocate(trash_wheel) |> 
  mutate(year = as.integer(year)) |> 
  mutate(plastic_bottles = as.numeric(plastic_bottles))

gwynnda_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                      sheet = "Gwynnda the Good Wheel of the W",
                      skip = 1,
                      range = cell_cols("A:L")) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(trash_wheel = "Gwynnda") |> 
  dplyr::relocate(trash_wheel) |> 
  mutate(year = as.integer(year))

trash_wheel_df = bind_rows(mrtrashwheel_df, proftrashwheel_df, gwynnda_df)
```

The combined Trash Wheel dataset has `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables, drawn from Mr. Trash Wheel
(`r sum(trash_wheel_df$trash_wheel == "MrTrashWheel")` hauls), Professor Trash Wheel
(`r sum(trash_wheel_df$trash_wheel == "ProfTrashWheel")` hauls), and Gwynnda
(`r sum(trash_wheel_df$trash_wheel == "Gwynnda")` hauls). Key variables include `trash_wheel`
(which wheel collected the trash), `dumpster`, `date`, `weight_tons`,
`volume_cubic_yards`, and counts of items such as `plastic_bottles`,
`cigarette_butts`, `wrappers`, and `sports_balls`. Sports balls and glass bottles
are only recorded for Mr. Trash Wheel, so they are missing for the other two.

Professor Trash Wheel collected a total of `r sum(proftrashwheel_df$weight_tons)` tons of trash.

Gwynnda collected `r gwynnda_df |> filter(month == "June", year == "2022") |> pull(cigarette_butts) |> sum()` cigarette butts in June 2022.
















